---
title: "Lecture 7 - An Introduction to R"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to set up R and RStudio on your machine. We will also demonstrate
  how to install R packages from CRAN, and install the tidyverse package.
---

```{r setup, include=FALSE}
library(learnr)
library(Brq) 
data("Prostate")
library(MixCluster) 
data("saheart") 
saheart = as.data.frame(saheart)
library(tidyverse)
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
```

## Lecture Objectives
 
- Learn how to: import, visualize and organize data.
 
### Outline

* Importing and or Loading Data
      * Setting Working Directory;
      * Reading Comma and Delimited Data Files.
  
* Data Visualization: An Introduction to `ggplot2`.
    * Layers;
    * Statistical transformations.
    
* Data Transformation: An Introduction to `dplyr`.
    * Filter rows;
    * Arrange rows;
    * Select columns;
    * Add new variables;
    * Grouped summaries.
  

## Importing and or Loading Data

### Setting Working

- Before reading any data, you must set the R working directory to the location of the data.

- `setwd(“…”)` will set the current working directory to a specific location

- `getwd()` will print out the current directory.

- `> setwd("C:/mydata")` (Windows)

- When specifying the pathname, R reads forward slashes, whereas Windows reads backward slashes. Definition of the path can be found at [path](https://en.wikipedia.org/wiki/Path_(computing))

- **Setting the working directory properly can eliminate path confusion.**

Source: [SAMHDA](https://www.datafiles.samhsa.gov/faq/how-do-i-read-data-r-nid3445)


### Reading Delimited Data Files 

Space-Delimited or Tab - Delimited

Function: `read.table()`

Common Parameters:

- File: the name of the file which the data are to be read from.

- Header: TRUE when first row includes variable names. The default is FALSE.

- Sep: A string indicating what is separating the data. The default is "  ".


'> dataSPACE <-read.table("C:/mydata/survey.dat", header=TRUE, sep= " ")'

or

'> dataSPACE <-read.table("C:/mydata/survey.txt", header=TRUE, sep= " ")'

### Reading comma delimited files

- `read.csv()` reads comma delimited files, 

- `read.csv2()` reads semicolon separated files (common in countries where , is used as the decimal place).

- However, we are going to use tidyverse packages to import data.


###Data import with Tidyverse

Most of readr’s functions are concerned with turning flat files into data frames:

- `read_csv()` reads comma delimited files, 

- `read_csv2()` reads semicolon separated files (common in countries where , is used as the decimal place), 

- `read_tsv()` reads tab delimited files, and `read_delim()` reads in files with any delimiter.

- There are other functions such as: `read_fwf()`, `read_log()`, however we are going focus on `read_csv()`.

###Example

```{r, warning=FALSE, message=FALSE}
x = read_csv("teste.csv") 
head(x)
```


### Test your knowledge

```{r quiz2, echo=FALSE}
quiz(caption = "Quiz - Importing and or Loading Data",
  question("What command do you use to set the working directory?",
    answer("setwd()", correct = TRUE, message = "Congrats!"),
    answer("library()", message="library and require load and attach add-on packages"),
    answer("getwd()", message="getwd returns an absolute filepath representing the current working directory of the R process"), allow_retry = TRUE
  ),
  question("What command do you use to read space delimited files?",
    answer("read.tsv()"),
    answer("read.table()", correct = TRUE, message = "Congrats!"),
    answer("read.csv()"),
    allow_retry = TRUE
  ),
  question("What command do you use to import comma delimited files?",
    answer("read.csv()", correct=TRUE, message = "You can also read comma delimited files using the function read_csv() or read.table() with sep=',' "),
    answer("read_tsv()"),
    answer("read.spss()" ),
    allow_retry = TRUE
  )
)
```

## Data Visualization:

### How to visualize data using R

This section is based on R for Data Science, section 5.6.

R has several systems for making graphs, but `ggplot2` is one of the most elegant, efficient and versatile.

There are many ways to make a data visualization, here we are going to focus on 4 basic representation types:

- Relationship.
- Comparison;
- Composition;
- Distribution.

Source: [Rbloggers](https://www.r-bloggers.com/7-visualizations-you-should-learn-in-r/)

### How to visualize relationship between variables: Scatter Plot

We use the scatter Plot to see the relationship between two continuous variables. 

Continuous variables are numeric variables that can assume values on an uncountable set.

####Example  - Prostate Cancer

The data came from a study by Stamey et al. (1989). They examined the correlation between the level of prostate-specific antigen and a number of clinical measures in men who were about to receive a radical prostatectomy. 

Here, we are going to analyze **the relationship between the logarithm of prostate weight and logarithm of prostate-specific antigen..**


### Scatter Plot - Prostate Cancer


```{r}
ggplot(Prostate, aes(x=lweight, y=lpsa)) + geom_point() + 
  scale_x_continuous("The logarithm of prostate weight", breaks = seq(0,6,by = 0.5))+     
  scale_y_continuous("The logarithm of prostate-specific antigen.", breaks = seq(0,6,by = 0.5))+
  theme_bw()

```

### Scatter Plot - Prostate Cancer (colors according to age)


```{r}
ggplot(Prostate, aes(x=lweight, y=lpsa, color=age)) + geom_point() + 
  scale_x_continuous("The logarithm of prostate weight", breaks = seq(0,6,by = 0.5))+     
  scale_y_continuous("The logarithm of prostate-specific antigen.", breaks = seq(0,6,by = 0.5))+  theme(legend.position="top")
```


### How to visualize the frequency distribution of a continuous variables: Histogram

The histogram consists of breaks the data into bins and computes the frequency distribution of them. We can modify the bin size and see what happens. 

There are some rules to define the number of bins available in the literature. 

By default, `geom_histogram()` will divide your data into 30 equal bins or intervals.

However, usually, the solution is ad hoc based on the experience of the statistician. Here, we are going to use the Sturges's Rule [references](https://en.wikipedia.org/wiki/Histogram#Sturges'_formula.)



### Histogram - Logarithm of prostate-specific antigen


```{r}

nbins = nclass.Sturges(Prostate$lpsa)

ggplot(Prostate, aes(lpsa)) + geom_histogram(bins=nbins, aes(y=..density..))+labs(title = "Histogram - log PSA") +xlab("The logarithm of prostate-specific antigen.") + theme_bw()
```


### How to visualize the frequency distribution of a discrete / categorical variables: Bar plot

####Exemple - South African Heart Disease

The data is a subset of the Coronary Risk-Factor Study (CORIS) baseline survey, carried
out in three rural areas of the Western Cape, South Africa (Rousseauw et
al., 1983). 

The aim of the study was to establish the intensity of ischemic
heart disease risk factors in that high-incidence region. 

Here we are going to investigate the frequency distribution of the variable coronary heart disease (CHD).

```{r, warning = FALSE, message=FALSE}
#install.packages("MCMCpack")
#install.packages("tmvtnorm")
#install.packages("MixCluster",repos="http://R-Forge.R-project.org")


ggplot(data=saheart, aes(x=CHD)) +
  geom_bar(aes(y = ..count../ sum(..count..))) + labs(title="Bar Plot") +
  ylab("Relative Frequency")



```

More details can be found at [STDHA](http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization).


### Stacked Bar chart: Analyzing association between the presence or not of heart disease in the family history and CHD clinic outcome

```{r, warning = FALSE, message=FALSE}

saheart %>% mutate(saheart,
  Famhist = factor(Famhist),
  CHD = factor(CHD)
)

ggplot(data=saheart, aes(x=CHD, fill = Famhist)) +
  geom_bar() + 
  labs(title="Stacked Bar Plot", x = "CHD", y="Absolute Frequency") +
  scale_fill_discrete(name = "Family history", labels = c("0", "1"))


```


### How to visualize the spread of the data and detect outlier: Box Plot

We use Box Plot to visualize the spread of the data and detect outliers. Also, we can use the Box Plot to make comparisons. Here we are going to compare the Obesity between groups with the presence of the CHD against the group with individuals who never had CHD.

```{r, warning = FALSE, message=FALSE}

ggplot(saheart, aes(CHD, Obesity)) + geom_boxplot(fill = "red")+
scale_y_continuous("Obesity Index", breaks= seq(0,50, by=5))+
labs(title = "Box Plot", x = "CHD")


```


## Data Transformation

This section is strongly based on [_R for Data Science_](http://r4ds.had.co.nz/)

### Filter rows

`filter()` lets you use a logical test to extract specific rows from a data frame. To use `filter()`, pass it the data frame followed by one or more logical tests. `filter()` will return every row that passes each logical test.

So we can use `filter()` to select every individual which has a coronary heart disease. Click Submit Answer to give it a try:

```{r filter1, exercise = TRUE, exercise.eval = FALSE}
filter(saheart, CHD == 1)
```

    
### Arrange rows

`arrange()` works similarly to `filter()` except that instead of selecting rows, it changes their order. It takes a data frame and a set of column names (or more complicated expressions) to order by. If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns:

So we can use `arrange()` to order the Prostate dataset by lcavol, lweight ,and age. Click Submit Answer to give it a try:

```{r arrange1, exercise = TRUE, exercise.eval = FALSE}
arrange(Prostate, lcavol, lweight, age)
```
  
Use desc to re-order by a column in descending order:  
   
```{r arrange2, exercise = TRUE, exercise.eval = FALSE}
arrange(Prostate,  desc(age))
```   
    


    
### Select columns

`select()` allows you to rapidly zoom in on a useful subset using operations based on the names of the variables.

For example, we can use `select()` to select the columns lcavol, lweight, age

```{r select1, exercise = TRUE, exercise.eval = FALSE}
select(Prostate, lcavol, lweight, age)
```
  
The function `select()` have many arguments:

everything(): Matches all variables.

last_col(): Select last variable, possibly with an offset.

These helpers select variables by matching patterns in their names:

- starts_with(): Starts with a prefix.

- ends_with(): Ends with a suffix.

- contains(): Contains a literal string.

- matches(): Matches a regular expression.

Further details can be found at documentation of the function.


### Add new variables

Besides selecting sets of existing columns, it’s often useful to add new columns that are functions of existing columns. That’s the job of mutate().

We can use `mutate()` to compute the level of the prostate-specific antigen.

```{r mutate1, exercise = TRUE, exercise.eval = FALSE}
mutate(Prostate, psa = exp(lpsa))
```
    
### Grouped summaries

`summarise()` collapses a data frame to a single row of summaries. You get to choose how many summaries appear in the row and how they are computed:

`summarise()` is not terribly useful unless you pair it with `group_by()`.

`group_by()` assigns groups by grouping together observations that have the same combinations of values for the variables that you pass to `group_by()`.


####Exemple 1 - Compute the average and sum of the lpsa for the entire data set.

```{r summarise1, exercise = TRUE, exercise.eval = FALSE}

Prostate %>%  summarise( lpsa_ave = mean(lpsa, na.rm=TRUE), 
                         lpsa_sum = sum(lpsa, na.rm=TRUE))

```


####Exemple 2 - Compute the average and sum of all variables for the entire data set.

```{r summarise2, exercise = TRUE, exercise.eval = FALSE}

Prostate %>% summarise_all(list(sum, mean))

```

####Exemple 3 - Compute the average for low-density lipoprotein cholesterol grouped by coronary heart disease event.

```{r summarise_groupedby1, exercise = TRUE, exercise.eval = FALSE, warning=FALSE, message=FALSE}

saheart %>% group_by(CHD) %>% summarise(mean(Ldl), n=n())

```

####Exemple 4 - Compute the absolute frequency for family history and  coronary heart disease event.

```{r summarise_groupedby2, exercise = TRUE, exercise.eval = FALSE,warning=FALSE, message=FALSE}

saheart %>% group_by(CHD,Famhist) %>% summarise(n())

```


Further details can be found at documentation.

##Exercises

To practice these skills, we will use the `flights` data set from the nycflights13 package. This data frame comes from the US [Bureau of Transportation Statistics](http://www.transtats.bts.gov/DatabaseInfo.asp?DB_ID=120&Link=0) and contains all `r format(nrow(nycflights13::flights), big.mark = ",")` flights that departed from New York City in 2013. It is documented in `?flights`.

Sort flights to find the most delayed flights.

```{r}
library(nycflights13)
```

```{r exerciseA, exercise = TRUE}


```

```{r exerciseA-solution}

arrange(flights, -arr_delay) 

```

Find the flights that left earliest.

```{r exerciseB, exercise = TRUE}


```

```{r exerciseB-solution}

arrange(flights, dep_delay) 

```

Add a new variable called `speed` (in seconds) in the dataset `flight`.

```{r exerciseC, exercise = TRUE}


```

```{r exerciseC-solution}
flights %>% mutate(speed = distance / (air_time * 60))

```

When do the first and last flights leave each day?
You must use the object `not_cancelled` to do th exercise.

```{r}
not_cancelled = flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))
```

```{r exerciseD, exercise = TRUE}


```

```{r exerciseD-solution}

 not_cancelled %>%
  group_by(year, month, day) %>% 
  summarise(
    first = min(dep_time),
    last = max(dep_time)
  )

```