---
title: "Lecture 20 - Assignment 2"
author: "Presented by Renato Rodrigues Silva"
institute: "Federal University of Goias."
date: "(updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false


---
class: middle
##Example - Student Performance Data Set

###Data Set Information:

- This dataset describes the student performance in secondary education of two Portuguese schools. 

- Two datasets are provided regarding the performance in two distinct subjects: Mathematics (mat) and Portuguese language (por). In [Cortez and Silva, 2008]

- Here, we are going to use only the dataset regarded to Portuguese

- The dataset was modeled under binary/five-level classification task.

- **Important note**: the target attribute G3 has a strong correlation with attributes G2 and G1. 

- This occurs because G3 is the final year grade (issued at the 3rd period), while G1 and G2 correspond to the 1st and 2nd period grades.

- It is more difficult to predict G3 without G2 and G1, but such prediction is much more useful (see paper source for more details).

- Here, we are going to consider G1, G2 and G3 as dichotomous variables.


---
class: middle
###Attribute Information:

- **school** - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)
- **sex** - student's sex (binary: 'F' - female or 'M' - male)
- **age** - student's age (numeric: from 15 to 22)
- **address** - student's home address type (binary: 'U' - urban or 'R' - rural)
- **famsize** - family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3)
- **Pstatus** - parent's cohabitation status (binary: 'T' - living together or 'A' - apart)
- **Medu** - mother's education (numeric: 0 - none, 1 - primary education (4th grade), 2 - 5th to 9th grade, 3 - secondary education, 4 - higher education)
- **Fedu** - father's education  (numeric: 0 - none, 1 - primary education (4th grade), 2 - 5th to 9th grade, 3 - secondary education, 4 - higher education)
- **Mjob** - mother's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')
- **Fjob** - father's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')

---
class: middle
###Attribute Information:


- **reason** - reason to choose this school (nominal: close to 'home', school 'reputation', 'course' preference or 'other')
- **guardian** - student's guardian (nominal: 'mother', 'father' or 'other')
- **traveltime** - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)
- **studytime** - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)
- **failures** - number of past class failures (numeric: n if 1<=n<3, else 4)
- **schoolsup** - extra educational support (binary: yes or no)
- **famsup** - family educational support (binary: yes or no)
- **paid** - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no)
- **activities** - extra-curricular activities (binary: yes or no)
- **nursery** - attended nursery school (binary: yes or no)


---
class: middle
###Attribute Information:


- **higher** - wants to take higher education (binary: yes or no)
- **internet** - Internet access at home (binary: yes or no)
- **romantic** - with a romantic relationship (binary: yes or no)
- **famrel** - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)
- **freetime** - free time after school (numeric: from 1 - very low to 5 - very high)
- **goout** - going out with friends (numeric: from 1 - very low to 5 - very high)
- **Dalc** - workday alcohol consumption (numeric: from 1 - very low to 5 - very high)
- **Walc** - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)
- **health** - current health status (numeric: from 1 - very bad to 5 - very good)
- **absences** - number of school absences (numeric: from 0 to 93)

---
class: middle
###Outcome:

- **G1** - first period grade (numeric: from 0 to 20)

- **G2** - second period grade (numeric: from 0 to 20)

- **G3** - final grade (numeric: from 0 to 20, output target)

- We are going to format the outcome as follows:

- $0 \leq G < 10 \Rightarrow \mbox{Failure};$

- $10 \leq G < 20  \Rightarrow \mbox{Approved};$

---
class: middle
###Logistic regression


```{r,warning = FALSE, message = FALSE, echo=FALSE}
library(tidyverse)
library(glmnet)
library(nnet)
library(caret)
library(ROCit)

dat =  read.csv2("student-por.csv", header = TRUE)



dat = mutate(dat,
            G1_final = ifelse(G1 <= 10, 0, 1
            ),
            G2_final =  ifelse(G2 <= 10, 0, 1
            ),
            G3_final = ifelse(G3 <= 10, 0, 1
            )) %>% select(-G1) %>% select(-G2) %>% select(-G3) 

names(dat)


```


---
class: middle
###Logistic regression

```{r,warning = FALSE, message = FALSE, echo=FALSE}


set.seed(42)
dat_idx = sample(nrow(dat), round(0.7*nrow(dat)))
dat_trn = dat[dat_idx, ]
dat_tst = dat[-dat_idx, ]


full.model = glm(G3_final~ ., data = dat_trn, 
                 family="binomial")


backward.model = MASS::stepAIC(full.model,
                               scope = list(upper = ~.,  lower = ~1),
                              family="binomial",
                              direction="backward",trace = FALSE)

summary(backward.model)

```


---
class: middle
###Logistic regression

```{r,warning = FALSE, message = FALSE, echo=FALSE}
#Testing data
score2 = predict(backward.model, type = "response",newdata = dat_tst) 
obs2 = dat_tst$G3_final
## Warning: package 'ROCit' was built under R version 3.5.2
ROCit_obj2 <- rocit(score=score2,class=obs2)
plot(ROCit_obj2)


```


---
class middle
##Example - Student Performance Data Set

```{r,warning = FALSE, message = FALSE, eval=FALSE}

#Testing data
optimal_cutoff = ROCit_obj2$Cutoff[which.max(ROCit_obj2$TPR - ROCit_obj2$FPR)]
model_glm_pred= ifelse(predict(backward.model, type = "response",newdata=dat_tst) > optimal_cutoff, "1", "0")
train_tab = table(predicted = model_glm_pred, actual = as.character(dat_tst$isBurned))
train_con_mat = confusionMatrix(train_tab, positive = "1")




```

---
class: middle
##Example - Student Performance Data Set

####Confusion Matrix
```{r, warning = FALSE, message = FALSE, eval=FALSE}
train_con_mat$table
```

####Accuracy
```{r, warning = FALSE, message = FALSE, eval=FALSE}
train_con_mat$overall["Accuracy"]
```

####Overall
```{r, warning = FALSE, message = FALSE, eval=FALSE}
train_con_mat$byClass
```


---
class: middle
##Example - Student Performance Data Set

```{r,warning = FALSE, message = FALSE, echo =FALSE}

optimal_cutoff = ROCit_obj2$Cutoff[which.max(ROCit_obj2$TPR - ROCit_obj2$FPR)]
model_glm_pred= ifelse(predict(backward.model, type = "response",newdata=dat_tst) > optimal_cutoff, "1", "0")
train_tab = table(predicted = model_glm_pred, actual = as.character(dat_tst$G3_final))
train_con_mat = confusionMatrix(train_tab, positive = "1")


```

####Optimal cutoff = `r optimal_cutoff`


####Confusion Matrix 
```{r, warning = FALSE, message = FALSE, echo =FALSE}
train_con_mat$table
```

####Accuracy = `r train_con_mat$overall["Accuracy"]`


####Overall
```{r, warning = FALSE, message = FALSE, echo=FALSE}
train_con_mat$byClass
```
