---
title: "Lecture 25 - torch for R"
author: "Material written by Daniel Falbel and Javier Luraschi. Presented by Renato Rodrigues Silva"
institute: "Federal University of Goias."
date: "(updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false


---
class: middle

##Installation

-   The package torch can be installed from CRAN with:

```{r, eval = FALSE, warning = FALSE, message = FALSE}
install.packages("torch")
```

- Alternatively, the development version of the  package also can be installed  with:

```{r, eval = FALSE, warning = FALSE, message = FALSE}
remotes::install_github("mlverse/torch")
```


---
class: middle

##Warm-up

###Example


-   A fully-connected ReLU network with one hidden layer and no biases, trained to predict y from x using mean squared error as cost function.

- In other words, ReLU function was used as activation function for hidden layer and linear function as activation function for output layer.

-   This implementation uses pure R to manually compute the forward pass, loss, and backward pass.

-   An R array is a generic n-dimensional array; it does not know anything about deep learning or gradients or computational graphs, and is just a way to perform generic numeric computations.


---
class: middle


####Forward pass  - For each instance we have:

\begin{align}
\mathbf{Z} =& \mathbf{X}_{n \times p}\mathbf{W}_{1 \phantom{1} p \times q} \\
\mathbf{A} =& g(\mathbf{Z})_{n \times q}    \\
\hat{\mathbf{y}} =&  \mathbf{A}_{n \times q}  \mathbf{W}_{2 \phantom{1} q \times 1} .
\end{align}


####Cost Function

$$L = (\mathbf{y} - \hat{\mathbf{y}})^{'}(\mathbf{y} - \hat{\mathbf{y}}) = (  \hat{\mathbf{y}} - \mathbf{y})^{'}( \hat{\mathbf{y}} - \mathbf{y} ) = \sum_{i=1}^n(y_i - \hat{y}_i)^2.$$

####Backpropagation ("Vectorized")


\begin{align}
\frac{\partial L}{\partial \mathbf{W}_2} = \frac{\partial L}{\partial \hat{\mathbf{y}}} \frac{\partial \hat{\mathbf{y}}}{\partial \mathbf{W}_2} = 2 \mathbf{A}^{'}(\hat{\mathbf{y}} - \mathbf{y}). 
\end{align}

\begin{align}
\frac{\partial L}{\partial \mathbf{W}_1} =\frac{\partial L}{\partial \hat{\mathbf{y}}} \frac{\partial \hat{\mathbf{y}}}{\partial \mathbf{A}} \frac{\partial \mathbf{A}}{\partial \mathbf{Z}} \frac{\partial \mathbf{Z}}{\partial \mathbf{W}_1} =& 2\mathbf{X}^{'} \frac{\partial g^{'}(\mathbf{Z})}{\partial \mathbf{Z}} \left[\left(\hat{\mathbf{y}} - \mathbf{y}\right)\mathbf{W}_2^{'} \right]  \\
=& 
2\mathbf{X}^{'}\left[\left(\hat{\mathbf{y}} - \mathbf{y}\right)\mathbf{W}_2^{'} \right]. 
\end{align}

[Source:](https://www.math.uwaterloo.ca/~hwolkowi/matrixcookbook.pdf)

---
class: middle

##Warm-up

- R code to simulate data and Randomly initialize weights

```{r, eval= FALSE, warning = FALSE, message = FALSE}

# N is batch size; D_in is input dimension;
# H is hidden dimension; D_out is output dimension.
N <- 64
D_in <- 1000
H <- 100
D_out <- 1

# Create random input and output data
x <- array(rnorm(N*D_in), dim = c(N, D_in))
y <- array(rnorm(N*D_out), dim = c(N, D_out))

# Randomly initialize weights
w1 <- array(rnorm(D_in*H), dim = c(D_in, H))
w2 <- array(rnorm(H*D_out), dim = c(H, D_out))


```


---
class:

##Warm-up

```{r, eval= FALSE, warning = FALSE, message = FALSE}
learning_rate <- 1e-6
for (t in seq_len(500)) {
   # Forward pass: compute predicted y
   h <- x %*% w1
   h_relu <- ifelse(h < 0, 0, h)
   y_pred <- h_relu %*% w2
   # Compute and print loss
   loss <- sum((y_pred - y)^2)
   if (t %% 100 == 0 || t == 1)
      cat("Step:", t, ":", loss, "\n")
   # Backprop to compute gradients of w1 and w2 with respect to loss
   grad_y_pred <- 2 * (y_pred - y)
   grad_w2 <- t(h_relu) %*% grad_y_pred
   grad_h_relu <- grad_y_pred %*% t(w2)
   grad_h <- grad_h_relu
   grad_h[h < 0] <- 0
   grad_w1 <- t(x) %*% grad_h
   # Update weights
   w1 <- w1 - learning_rate * grad_w1
   w2 <- w2 - learning_rate * grad_w2
}


```


---
class: middle

##Warm-up

- The output of the R code as follows

```{r, echo= FALSE, warning = FALSE, message = FALSE}

# N is batch size; D_in is input dimension;
# H is hidden dimension; D_out is output dimension.
N <- 64
D_in <- 1000
H <- 100
D_out <- 1

# Create random input and output data
x <- array(rnorm(N*D_in), dim = c(N, D_in))
y <- array(rnorm(N*D_out), dim = c(N, D_out))

# Randomly initialize weights
w1 <- array(rnorm(D_in*H), dim = c(D_in, H))
w2 <- array(rnorm(H*D_out), dim = c(H, D_out))

learning_rate <- 1e-6
for (t in seq_len(500)) {
   # Forward pass: compute predicted y
   h <- x %*% w1
   h_relu <- ifelse(h < 0, 0, h)
   y_pred <- h_relu %*% w2
   
   # Compute and print loss
   loss <- sum((y_pred - y)^2)
   if (t %% 100 == 0 || t == 1)
      cat("Step:", t, ":", loss, "\n")
   
   # Backprop to compute gradients of w1 and w2 with respect to loss
   grad_y_pred <- 2 * (y_pred - y)
   grad_w2 <- t(h_relu) %*% grad_y_pred
   grad_h_relu <- grad_y_pred %*% t(w2)
   grad_h <- grad_h_relu
   grad_h[h < 0] <- 0
   grad_w1 <- t(x) %*% grad_h
   
   # Update weights
   w1 <- w1 - learning_rate * grad_w1
   w2 <- w2 - learning_rate * grad_w2
}


```

- Now, we will replace the R array with a torch Tensor.

- However, we need to learn what is torch Tensor first.

---
class: middle

##Tensors

-    A tensor is a multidimensional array, in other words, tensors is a general mathematical structure that could have more than two dimensions.

###Torch Tensor

-   Unfortunately, R arrays cannot utilize GPUs to accelerate its numerical computations. For modern deep neural networks, GPUs often provide speedups of 50x or greater.

- To deal with this problem, we can use the torch tensor. A torch tensor is an n-dimensional array, which has many functions for mathematical /computational operations. 

- Unlike R, torch Tensors can utilize GPUs to accelerate their numeric computations. To run a torch Tensor on GPU, you simply need to cast it to a new datatype.



---
class: middle

##Creating tensors from R objects

-   The torch_tensor function takes an R vector, matrix or array and creates an equivalent torch_tensor.

```{r, eval = FALSE, message = FALSE, warning = FALSE}

library(torch)

torch_tensor(c(1,2,3))

torch_tensor(matrix(1:10, ncol = 5, nrow = 2, byrow = TRUE))

torch_tensor(array(runif(12), dim = c(2, 2, 3)))


```

- By default, we will create tensors in the cpu device, converting their R datatype to the corresponding torch dtype.

- Note currently, only numeric and boolean types are supported.

---
class: middle


```{r, echo = FALSE, message = FALSE, warning = FALSE}

library(torch)

torch_tensor(c(1,2,3))

torch_tensor(matrix(1:10, ncol = 5, nrow = 2, byrow = TRUE))

torch_tensor(array(runif(12), dim = c(2, 2, 3)))


```

---
class: middle

##Creating tensors from R objects

-   You can always modify dtype and device when converting an R object to a torch tensor. For example:

```{r, message = FALSE, warning = FALSE}

torch_tensor(1, dtype = torch_long())

torch_tensor(1, device = "cpu", dtype = torch_float64())

```

---
class: middle

##Using creation functions

Here is the full list of functions that can be used to bulk-create tensors in torch:

-   torch_arange: Returns a tensor with a sequence of integers,
-   torch_empty: Returns a tensor with uninitialized values,
-   torch_eye: Returns an identity matrix,
-   torch_full: Returns a tensor filled with a single value,
-   torch_linspace: Returns a tensor with values linearly spaced in some interval,
-   torch_logspace: Returns a tensor with values logarithmically spaced in some interval,
-   torch_ones: Returns a tensor filled with all ones,
-   torch_rand: Returns a tensor filled with values drawn from a uniform distribution on $[0, 1)$.
-   torch_randint: Returns a tensor with integers randomly drawn from an interval,
-   torch_randn: Returns a tensor filled with values drawn from a unit normal distribution,
-   torch_randperm: Returns a tensor filled with a random permutation of integers in some interval,
-   torch_zeros: Returns a tensor filled with all zeros.


---
class: middle

##Using creation functions - Examples

```{r, message = FALSE, warning = FALSE}
x <- torch_randn(5, 3)
x

x <- torch_ones(2, 4, dtype = torch_int64(), device = "cpu")
x

```

---
class: middle

##Conversion - Example

```{r, message = FALSE, warning = FALSE}

x <- torch_tensor(1)
x

y <- x$to(dtype = torch_int32())
y


```


-   You can also copy a tensor to the GPU using:

```{r, eval= FALSE, message = FALSE, warning = FALSE}
#NOT RUN!!!
x <- torch_tensor(1)

y <- x$cuda()
y

```


---
class: middle

##Tensor objects and methods

-   Tensors have a large amount of methods that can be called using the $ operator.

-   A list of all methods that can be called by tensor objects and their documentation can be found [here.](https://torch.mlverse.org/docs/articles/tensor/index.html.)

- You can also look at PyTorch’s documentation for additional details.



---
class: middle

##Neural Network revisited using the tensors

```{r, eval= FALSE, message = FALSE, warning = FALSE}

if (cuda_is_available()) {
   device <- torch_device("cuda")
} else {
   device <- torch_device("cpu")
}
   
# N is batch size; D_in is input dimension;
# H is hidden dimension; D_out is output dimension.
N <- 64
D_in <- 1000
H <- 100
D_out <- 1

# Create random input and output data
x <- torch_randn(N, D_in, device=device)
y <- torch_randn(N, D_out, device=device)

# Randomly initialize weights
w1 <- torch_randn(D_in, H, device=device)
w2 <- torch_randn(H, D_out, device=device)

```



---
class: middle

##Neural Network revisited using the tensors

```{r, eval= FALSE, message = FALSE, warning = FALSE}

learning_rate <- 1e-6
for (t in seq_len(1000)) {
   # Forward pass: compute predicted y
   h <- x$mm(w1)
   h_relu <- h$clamp(min=0)
   y_pred <- h_relu$mm(w2)
   # Compute and print loss
   loss <- as.numeric((y_pred - y)$pow(2)$sum())
   if (t %% 100 == 0 || t == 1)
      cat("Step:", t, ":", loss, "\n")
   # Backprop to compute gradients of w1 and w2 with respect to loss
   grad_y_pred <- 2.0 * (y_pred - y)
   grad_w2 <- h_relu$t()$mm(grad_y_pred)
   grad_h_relu <- grad_y_pred$mm(w2$t())
   grad_h <- grad_h_relu$clone()
   grad_h[h < 0] <- 0
   grad_w1 <- x$t()$mm(grad_h)
   # Update weights using gradient descent
   w1 <- w1 - learning_rate * grad_w1
   w2 <- w2 - learning_rate * grad_w2
}

```




---
class: middle

##Neural Network revisited using the tensors


- The output of R code as follows:

```{r, echo= FALSE, message = FALSE, warning = FALSE}

if (cuda_is_available()) {
   device <- torch_device("cuda")
} else {
   device <- torch_device("cpu")
}
   
# N is batch size; D_in is input dimension;
# H is hidden dimension; D_out is output dimension.
N <- 64
D_in <- 1000
H <- 100
D_out <- 1

# Create random input and output data
x <- torch_randn(N, D_in, device=device)
y <- torch_randn(N, D_out, device=device)

# Randomly initialize weights
w1 <- torch_randn(D_in, H, device=device)
w2 <- torch_randn(H, D_out, device=device)


learning_rate <- 1e-6
for (t in seq_len(1000)) {
   # Forward pass: compute predicted y
   h <- x$mm(w1)
   h_relu <- h$clamp(min=0)
   y_pred <- h_relu$mm(w2)
   # Compute and print loss
   loss <- as.numeric((y_pred - y)$pow(2)$sum())
   if (t %% 100 == 0 || t == 1)
      cat("Step:", t, ":", loss, "\n")
   # Backprop to compute gradients of w1 and w2 with respect to loss
   grad_y_pred <- 2.0 * (y_pred - y)
   grad_w2 <- h_relu$t()$mm(grad_y_pred)
   grad_h_relu <- grad_y_pred$mm(w2$t())
   grad_h <- grad_h_relu$clone()
   grad_h[h < 0] <- 0
   grad_w1 <- x$t()$mm(grad_h)
   # Update weights using gradient descent
   w1 <- w1 - learning_rate * grad_w1
   w2 <- w2 - learning_rate * grad_w2
}

```

-   In the next example, we will use autograd instead of computing the gradients manually.

-   However, let's learn how to autograd works.


---
class: middle

##Tensors and Autograd


-   In the previous examples, we had to manually implement both the forward and backward passes of our neural network. 

-   Manually implementing the backward pass is not a big deal for a small two-layer network, but can quickly get very hairy for large complex networks.

-   We can use automatic differentiation to automate the computation of backward passes in neural networks. 

-   The autograd feature in torch provides exactly this functionality. 

-   When using autograd, the forward pass of your network will define a computational graph; nodes in the graph will be Tensors, and edges will be functions that produce output Tensors from input Tensors. 

-   Backpropagating through this graph then allows you to easily compute gradients.



---
class: middle

##Tensors and Autograd


-   Each Tensor represents a node in a computational graph. If x is a Tensor that has x$requires_grad=TRUE then x$grad is another Tensor holding the gradient of x with respect to some scalar value.

-   Here we use torch Tensors and autograd to implement our two-layer network; now we no longer need to manually implement the backward pass through the network:



---
class: middle

##Tensors and Autograd - Loading and Initialization


```{r, eval= FALSE, message = FALSE, warning = FALSE}
if (cuda_is_available()) {
   device <- torch_device("cuda")
} else {
   device <- torch_device("cpu")
}
# N is batch size; D_in is input dimension;
# H is hidden dimension; D_out is output dimension.
N <- 64
D_in <- 1000
H <- 100
D_out <- 1
# Create random input and output data
# Setting requires_grad=FALSE (the default) indicates that we do not need to 
# compute gradients with respect to these Tensors during the backward pass.
x <- torch_randn(N, D_in, device=device)
y <- torch_randn(N, D_out, device=device)
# Randomly initialize weights
# Setting requires_grad=TRUE indicates that we want to compute gradients with
# respect to these Tensors during the backward pass.
w1 <- torch_randn(D_in, H, device=device, requires_grad = TRUE)
w2 <- torch_randn(H, D_out, device=device, requires_grad = TRUE)
```


---
class: middle

##Tensors and Autograd - Forward pass



```{r, eval= FALSE, message = FALSE, warning = FALSE}

learning_rate <- 1e-6
for (t in seq_len(500)) {
   # Forward pass: compute predicted y using operations on Tensors; these
   # are exactly the same operations we used to compute the forward pass using
   # Tensors, but we do not need to keep references to intermediate values since
   # we are not implementing the backward pass by hand.
   y_pred <- x$mm(w1)$clamp(min=0)$mm(w2)
   
   # Compute and print loss using operations on Tensors.
   # Now loss is a Tensor of shape (1,)
   loss <- (y_pred - y)$pow(2)$sum()
   if (t %% 100 == 0 || t == 1)
      cat("Step:", t, ":", as.numeric(loss), "\n")


```



---
class: middle

##Tensors and Autograd - Backpropgation pass

```{r, eval= FALSE, message = FALSE, warning = FALSE}

# Use autograd to compute the backward pass. This call will compute the
   # gradient of loss with respect to all Tensors with requires_grad=True.
   # After this call w1$grad and w2$grad will be Tensors holding the gradient
   # of the loss with respect to w1 and w2 respectively.
   loss$backward()
   
   # Manually update weights using gradient descent. Wrap in `with_no_grad`
   # because weights have requires_grad=TRUE, but we don't need to track this
   # in autograd.
   # You can also use optim_sgd to achieve this.
   with_no_grad({
      
      # operations suffixed with an `_` operates on in-place on the tensor.
      w1$sub_(learning_rate * w1$grad)
      w2$sub_(learning_rate * w2$grad)
      
      # Manually zero the gradients after updating weights
      w1$grad$zero_()
      w2$grad$zero_()
   })
}

```




---
class: middle

##Tensors and Autograd 

- The output of the R code is the following

```{r, echo= FALSE, message = FALSE, warning = FALSE}

if (cuda_is_available()) {
   device <- torch_device("cuda")
} else {
   device <- torch_device("cpu")
}
   
# N is batch size; D_in is input dimension;
# H is hidden dimension; D_out is output dimension.
N <- 64
D_in <- 1000
H <- 100
D_out <- 1

# Create random input and output data
# Setting requires_grad=FALSE (the default) indicates that we do not need to 
# compute gradients with respect to these Tensors during the backward pass.
x <- torch_randn(N, D_in, device=device)
y <- torch_randn(N, D_out, device=device)

# Randomly initialize weights
# Setting requires_grad=TRUE indicates that we want to compute gradients with
# respect to these Tensors during the backward pass.
w1 <- torch_randn(D_in, H, device=device, requires_grad = TRUE)
w2 <- torch_randn(H, D_out, device=device, requires_grad = TRUE)

learning_rate <- 1e-6
for (t in seq_len(500)) {
   # Forward pass: compute predicted y using operations on Tensors; these
   # are exactly the same operations we used to compute the forward pass using
   # Tensors, but we do not need to keep references to intermediate values since
   # we are not implementing the backward pass by hand.
   y_pred <- x$mm(w1)$clamp(min=0)$mm(w2)
   
   # Compute and print loss using operations on Tensors.
   # Now loss is a Tensor of shape (1,)
   loss <- (y_pred - y)$pow(2)$sum()
   if (t %% 100 == 0 || t == 1)
      cat("Step:", t, ":", as.numeric(loss), "\n")
   
   # Use autograd to compute the backward pass. This call will compute the
   # gradient of loss with respect to all Tensors with requires_grad=True.
   # After this call w1$grad and w2$grad will be Tensors holding the gradient
   # of the loss with respect to w1 and w2 respectively.
   loss$backward()
   
   # Manually update weights using gradient descent. Wrap in `with_no_grad`
   # because weights have requires_grad=TRUE, but we don't need to track this
   # in autograd.
   # You can also use optim_sgd to achieve this.
   with_no_grad({
      
      # operations suffixed with an `_` operates on in-place on the tensor.
      w1$sub_(learning_rate * w1$grad)
      w2$sub_(learning_rate * w2$grad)
      
      # Manually zero the gradients after updating weights
      w1$grad$zero_()
      w2$grad$zero_()
   })
}

```

- Now, we will learn we will learn how to use the neural networks abstractions in torch.


---
class: middle

##nn: neural networks with torch


-   Computational graphs and autograd are a very powerful paradigm for defining complex operators and automatically taking derivatives; however for large neural networks raw autograd can be a bit too low-level.

-   When building neural networks we frequently think of arranging the computation into layers, some of which have learnable parameters which will be optimized during learning.

-   In torch, the nn functionality serves this same purpose. The nn feature defines a set of Modules, which are roughly equivalent to neural network layers. 

-   A Module receives input Tensors and computes output Tensors, but may also hold internal state such as Tensors containing learnable parameters. 

-   The nn collection also defines a set of useful loss functions that are commonly used when training neural networks.


---
class: middle


##nn: neural networks with torch


```{r, eval= FALSE, message = FALSE, warning = FALSE}

if (cuda_is_available()) {
   device <- torch_device("cuda")
} else {
   device <- torch_device("cpu")
}
   
# N is batch size; D_in is input dimension;
# H is hidden dimension; D_out is output dimension.
N <- 64
D_in <- 1000
H <- 100
D_out <- 1

# Create random input and output data
# Setting requires_grad=FALSE (the default) indicates that we do not need to 
# compute gradients with respect to these Tensors during the backward pass.
x <- torch_randn(N, D_in, device=device)
y <- torch_randn(N, D_out, device=device)

```



---
class: middle


##nn: neural networks with torch


```{r, eval= FALSE, message = FALSE, warning = FALSE}

# Use the nn package to define our model as a sequence of layers. nn_sequential
# is a Module which contains other Modules, and applies them in sequence to
# produce its output. Each Linear Module computes output from input using a
# linear function, and holds internal Tensors for its weight and bias.
model <- nn_sequential(
    nn_linear(D_in, H),
    nn_relu(),
    nn_linear(H, D_out)
)

# The nn package also contains definitions of popular loss functions; in this
# case we will use Mean Squared Error (MSE) as our loss function.
loss_fn <- nnf_mse_loss


```




---
class: middle


##nn: neural networks with torch


```{r, eval= FALSE, message = FALSE, warning = FALSE}

learning_rate <- 1e-6
for (t in seq_len(500)) {
   # Forward pass: compute predicted y by passing x to the model. Module objects
   # can be called like functions. When doing so you pass a Tensor of input
   # data to the Module and it produces a Tensor of output data.
   y_pred <- model(x)
   # Compute and print loss. We pass Tensors containing the predicted and true
   # values of y, and the loss function returns a Tensor containing the
   # loss.
   loss <- loss_fn(y_pred, y)
   if (t %% 100 == 0 || t == 1)
      cat("Step:", t, ":", as.numeric(loss), "\n")
   # Zero the gradients before running the backward pass.
   model$zero_grad()
   # Backward pass: compute gradient of the loss with respect to all the learnable
   # parameters of the model. Internally, the parameters of each Module are stored
   # in Tensors with requires_grad=TRUE, so this call will compute gradients for
   # all learnable parameters in the model.
   loss$backward()

   
``` 
     
---
class: middle

##nn: neural networks with torch


```{r, eval= FALSE, message = FALSE, warning = FALSE}  
   
   
   # Update the weights using gradient descent. Each parameter is a Tensor, so
   # we can access its gradients like we did before.
   with_no_grad({
      for (param in model$parameters) {
         param$sub_(learning_rate * param$grad)
      }
   })
}

```



---
class: middle

##nn: neural networks with torch

- The output of the R code is the following

```{r, echo= FALSE, message = FALSE, warning = FALSE}  

if (cuda_is_available()) {
   device <- torch_device("cuda")
} else {
   device <- torch_device("cpu")
}
   
# N is batch size; D_in is input dimension;
# H is hidden dimension; D_out is output dimension.
N <- 64
D_in <- 1000
H <- 100
D_out <- 1

# Create random input and output data
# Setting requires_grad=FALSE (the default) indicates that we do not need to 
# compute gradients with respect to these Tensors during the backward pass.
x <- torch_randn(N, D_in, device=device)
y <- torch_randn(N, D_out, device=device)

# Use the nn package to define our model as a sequence of layers. nn_sequential
# is a Module which contains other Modules, and applies them in sequence to
# produce its output. Each Linear Module computes output from input using a
# linear function, and holds internal Tensors for its weight and bias.
model <- nn_sequential(
    nn_linear(D_in, H),
    nn_relu(),
    nn_linear(H, D_out)
)

# The nn package also contains definitions of popular loss functions; in this
# case we will use Mean Squared Error (MSE) as our loss function.
loss_fn <- nnf_mse_loss

learning_rate <- 1e-6
for (t in seq_len(500)) {
   # Forward pass: compute predicted y by passing x to the model. Module objects
   # can be called like functions. When doing so you pass a Tensor of input
   # data to the Module and it produces a Tensor of output data.
   y_pred <- model(x)
   
   # Compute and print loss. We pass Tensors containing the predicted and true
   # values of y, and the loss function returns a Tensor containing the
   # loss.
   loss <- loss_fn(y_pred, y)
   if (t %% 100 == 0 || t == 1)
      cat("Step:", t, ":", as.numeric(loss), "\n")
   
   # Zero the gradients before running the backward pass.
   model$zero_grad()

   # Backward pass: compute gradient of the loss with respect to all the learnable
   # parameters of the model. Internally, the parameters of each Module are stored
   # in Tensors with requires_grad=TRUE, so this call will compute gradients for
   # all learnable parameters in the model.
   loss$backward()
   
   # Update the weights using gradient descent. Each parameter is a Tensor, so
   # we can access its gradients like we did before.
   with_no_grad({
      for (param in model$parameters) {
         param$sub_(learning_rate * param$grad)
      }
   })
}



```

-  Finally, we will learn how to use optimizers implemented in torch.


---
class: middle

##optim: optimizers in torch

- We often train neural networks using more sophisticated optimizers like AdaGrad, RMSProp, Adam, etc.

- The optim package in torch abstracts the idea of an optimization algorithm and provides implementations of commonly used optimization algorithms.

- In this example we will use the nn package to define our model as before, but we will optimize the model using the Adam algorithm provided by optim:


---
class: middle

##optim: optimizers in torch

```{r, eval = FALSE, message = FALSE, warning = FALSE}  

if (cuda_is_available()) {
   device <- torch_device("cuda")
} else {
   device <- torch_device("cpu")
}
   
# N is batch size; D_in is input dimension;
# H is hidden dimension; D_out is output dimension.
N <- 64
D_in <- 1000
H <- 100
D_out <- 10

# Create random input and output data
# Setting requires_grad=FALSE (the default) indicates that we do not need to 
# compute gradients with respect to these Tensors during the backward pass.
x <- torch_randn(N, D_in, device=device)
y <- torch_randn(N, D_out, device=device)


```


---
class: middle

##optim: optimizers in torch

```{r, eval = FALSE, message = FALSE, warning = FALSE}  

##optim: optimizers in torch

# Use the nn package to define our model as a sequence of layers. nn_sequential
# is a Module which contains other Modules, and applies them in sequence to
# produce its output. Each Linear Module computes output from input using a
# linear function, and holds internal Tensors for its weight and bias.
model <- nn_sequential(
    nn_linear(D_in, H),
    nn_relu(),
    nn_linear(H, D_out)
)

# The nn package also contains definitions of popular loss functions; in this
# case we will use Mean Squared Error (MSE) as our loss function.
loss_fn <- nnf_mse_loss

# Use the optim package to define an Optimizer that will update the weights of
# the model for us. Here we will use Adam; the optim package contains many other
# optimization algorithms. The first argument to the Adam constructor tells the
# optimizer which Tensors it should update.
learning_rate <- 1e-4
optimizer <- optim_adam(model$parameters, lr=learning_rate)



```




---
class: middle

###optim: optimizers in torch

```{r, eval = FALSE, message = FALSE, warning = FALSE}  

for (t in seq_len(500)) {
   # Forward pass: compute predicted y by passing x to the model. Module objects
   # can be called like functions. When doing so you pass a Tensor of input
   # data to the Module and it produces a Tensor of output data.
   y_pred <- model(x)
   # Compute and print loss. We pass Tensors containing the predicted and true
   # values of y, and the loss function returns a Tensor containing the
   # loss.
   loss <- loss_fn(y_pred, y)
   if (t %% 100 == 0 || t == 1)
      cat("Step:", t, ":", as.numeric(loss), "\n")
   # Before the backward pass, use the optimizer object to zero all of the
   # gradients for the variables it will update (which are the learnable
   # weights of the model). This is because by default, gradients are
   # accumulated in buffers( i.e, not overwritten) whenever $backward()
   # is called. Checkout docs of `autograd_backward` for more details.
   optimizer$zero_grad()
   # Backward pass: compute gradient of the loss with respect to model
   # parameters
   loss$backward()
   # Calling the step function on an Optimizer makes an update to its
   # parameters
   optimizer$step()
}


```



---
class: middle

###optim: optimizers in torch

- The output of the R code is the following

```{r, echo = FALSE, message = FALSE, warning = FALSE}  

if (cuda_is_available()) {
   device <- torch_device("cuda")
} else {
   device <- torch_device("cpu")
}
   
# N is batch size; D_in is input dimension;
# H is hidden dimension; D_out is output dimension.
N <- 64
D_in <- 1000
H <- 100
D_out <- 10

# Create random input and output data
# Setting requires_grad=FALSE (the default) indicates that we do not need to 
# compute gradients with respect to these Tensors during the backward pass.
x <- torch_randn(N, D_in, device=device)
y <- torch_randn(N, D_out, device=device)

# Use the nn package to define our model as a sequence of layers. nn_sequential
# is a Module which contains other Modules, and applies them in sequence to
# produce its output. Each Linear Module computes output from input using a
# linear function, and holds internal Tensors for its weight and bias.
model <- nn_sequential(
    nn_linear(D_in, H),
    nn_relu(),
    nn_linear(H, D_out)
)

# The nn package also contains definitions of popular loss functions; in this
# case we will use Mean Squared Error (MSE) as our loss function.
loss_fn <- nnf_mse_loss

# Use the optim package to define an Optimizer that will update the weights of
# the model for us. Here we will use Adam; the optim package contains many other
# optimization algorithms. The first argument to the Adam constructor tells the
# optimizer which Tensors it should update.
learning_rate <- 1e-4
optimizer <- optim_adam(model$parameters, lr=learning_rate)

for (t in seq_len(500)) {
   # Forward pass: compute predicted y by passing x to the model. Module objects
   # can be called like functions. When doing so you pass a Tensor of input
   # data to the Module and it produces a Tensor of output data.
   y_pred <- model(x)
   
   # Compute and print loss. We pass Tensors containing the predicted and true
   # values of y, and the loss function returns a Tensor containing the
   # loss.
   loss <- loss_fn(y_pred, y)
   if (t %% 100 == 0 || t == 1)
      cat("Step:", t, ":", as.numeric(loss), "\n")
   
   # Before the backward pass, use the optimizer object to zero all of the
   # gradients for the variables it will update (which are the learnable
   # weights of the model). This is because by default, gradients are
   # accumulated in buffers( i.e, not overwritten) whenever $backward()
   # is called. Checkout docs of `autograd_backward` for more details.
   optimizer$zero_grad()

   # Backward pass: compute gradient of the loss with respect to model
   # parameters
   loss$backward()

   # Calling the step function on an Optimizer makes an update to its
   # parameters
   optimizer$step()
}

```

- Other examples could be found 
[here](http://www.sharetechnote.com/html/Python_PyTorch_nn_Sequential_01.html) or in [in PyTorch documentation.](https://pytorch.org/docs/stable/nn.html#containers)
