---
title: "Lecture 8 - Simple Linear Regression part I"
author: " Presented by Renato Rodrigues Silva"
institute: "Federal University of Goias."
date: "(updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false


---
###Some advices before start (Julian J.Faraway)

- Do not begin a complex statistical analysis without observing the objectives, or even whether the data are appropriate for the model proposed.


###Formulation of the problem

a. Understand the "physical background".

b. Understand the objective.

c. Make sure you know what the client wants: Keep it as simple as possible.

d. Put the problem into statistical terms.



"The formulation of a problem is often more essential than its solution which
may be merely a matter of mathematical or experimental skill". (Albert Einstein)


---
##When to Use Regression Analysis (Julian J.Faraway)

-  Regression models are used for explaining or modeling the relationship between a single variable, called the response, output or dependent variable; and one or more predictors, inputs, or explanatory variables.

- The response variable must be a continuous, but the predictors can be
continuous, discrete or categorical.

- Simple Linear Regression is a case particular when we have only one predictor.


##Potential Goals of a Regression analysis:

1. Prediction of future observations;
2. Assessment of the effect of, or relationship between, explanatory variables and the
response;
3. A general description of data structure.


---
class: middle
##Simple Linear Regression - Intuiton


####Example  - Prostate Cancer

The data came from a study by Stamey et al. (1989). They examined the correlation between the level of prostate-specific antigen and a number of clinical measures in men who were about to receive a radical prostatectomy. 

The data frame has the following components:

- lcavol : The logarithm of cancer volume.
- lweight : The logarithm of prostate weight.
- age : Age
- lbph : The logarithm of the amount of benign prostatic hyperplasia.
- svi : Seminal vesicle invasion.
- lcp : The logarithm of capsular penetration.
- gleason : Gleason score.
- pgg45 : Percentage Gleason scores 4 or 5.
- lpsa : The logarithm of prostate-specific antigen.

Today, our objective is study **the relationship the logarithm of cancer volume and logarithm of prostate-specific antigen.**

---
### Scatter Plot - Prostate Cancer


```{r, warning=FALSE, message=FALSE,echo=FALSE}
library(Brq)
library(tidyverse)
library(broom)
data("Prostate")
ggplot(Prostate, aes(x=lcavol, y=lpsa)) + geom_point() + 
  #scale_x_continuous("The logarithm of cancer volume", breaks = seq(0,6,by = 0.5))+     
  #scale_y_continuous("The logarithm of prostate-specific antigen.", breaks = seq(0,6,by = 0.5))+
  theme_bw()

```


---
## Questions

- What is the relationship between the logarithm of cancer volume and the logarithm of prostate-specific antigen?
- Is it reasonable to assume a straight line?
- If yes, what are the values of the intercept and the slope?
- How to fit a straight line to the data?

##Simple Linear Regression

$$y = \beta_0 + \beta_1 x + \epsilon$$
where $y$ is the response variable, $x$ is the explanatory variable, $\epsilon_i$ is the random error.

The simple regression model is linear in the $\beta_0$ and $\beta_1$ parameters. However, it is not necessarily linear in the $x$ variables. Thus models such as $y = \beta_0 + \beta_1 \log{x} + \epsilon$ are still designed as simple linear regression.

---

##Simple Linear Regression for n observations

Given $(x_1, y_1), \ldots, (x_n, y_n)$, we have

$$y_i = \beta_0 +  \beta_1 x_i + \epsilon_i,$$

##Assumptions

1.  $E(\epsilon_i) = 0, \phantom{11} \forall i = 1, 2, \ldots, ...$
2.  $Var(\epsilon_i) = 0, \phantom{11} \forall i = 1, 2, \ldots, ...$
3.  $Cov(\epsilon_i,\epsilon_j) = 0, \phantom{11} \forall i \neq j$.
4.  $x_i$ are known constants

---
class: middle
##Estimation  

Let's get started this section watching this [video](https://www.youtube.com/watch?v=PaFPbb66DxQ&feature=youtu.be) 


###Least Square Approach

Considering the error sum of squares given by  $Q = \sum_{i=1}^n (y_i - \beta_0 - \beta_1 x_i)^2$, the least square solution is obtained solving the following minimization problem:

$$(\hat{\beta}_0, \hat{\beta}_1) = \mbox{arg min}_{(\beta_0, \beta_1)} \sum_{i=1}^n (y_i - \beta_0 - \beta_1 x_i)^2.$$

---

###How to obtain the estimators ?

Taking deritatives of $Q$ with respect to  $\beta_0$ and $\beta_1$, we have


--
 $$\frac{\partial Q}{\partial \beta_0} = -2\sum_{i=1}^n (y_i - \beta_0 - \beta_1 x_i) = -2\left[ \sum_{i=1}^n y_i - n \beta_0 - \beta_1 \sum_{i=1}^n x_i\right]$$

--

$$\frac{\partial Q}{\partial \beta_1} = -2\sum_{i=1}^n (y_i - \beta_0 - \beta_1 x_i)x_i = -2\left[ \sum_{i=1}^n y_ix_i -  \beta_0\sum_{i=1}^nx_i - \beta_1 \sum_{i=1}^n x_i^2\right]$$

Now, making $\frac{\partial Q}{\partial \beta_0} =0$ and $\frac{\partial Q}{\partial \beta_1} =0$, we obtain the normal equations, i.e,

\begin{align}
\sum_{i=1}^n y_i - n \beta_0 - \beta_1 \sum_{i=1}^n x_i =& 0 \\
\sum_{i=1}^n y_ix_i -  \beta_0\sum_{i=1}^nx_i - \beta_1 \sum_{i=1}^n x_i^2 =&0.
\end{align}

---

###Normal Equations

\begin{align}
\sum_{i=1}^n y_i - n \hat{\beta}_0 - \hat{\beta}_1 \sum_{i=1}^n x_i =& 0 \\
\sum_{i=1}^n y_ix_i -  \hat{\beta}_0\sum_{i=1}^nx_i - \hat{\beta}_1 \sum_{i=1}^n x_i^2 =&0.
\end{align}

####First line

--
$$n \hat{\beta}_0  = \sum_{i=1}^n y_i - \hat{\beta}_1 \sum_{i=1}^n x_i$$
--
$$\hat{\beta}_0 = \bar{y} -  \hat{\beta}_1 \bar{x}.$$

---
####Second line

--
$$\hat{\beta}_1 \sum_{i=1}^n x_i^2 = \sum_{i=1}^n y_ix_i -  \hat{\beta}_0\sum_{i=1}^nx_i$$
--
$$\hat{\beta}_1 \sum_{i=1}^n x_i^2 = \sum_{i=1}^n y_ix_i -  (\bar{y} -  \hat{\beta}_1 \bar{x})\sum_{i=1}^nx_i$$
--
$$\hat{\beta}_1 \sum_{i=1}^n x_i^2 -  \hat{\beta}_1 \bar{x}\sum_{i=1}^nx_i = \sum_{i=1}^n y_ix_i -  \bar{y}\sum_{i=1}^nx_i$$
--
$$\hat{\beta}_1 (\sum_{i=1}^n x_i^2 -   \frac{(\sum_{i=1}^nx_i)^2}{n}) = \sum_{i=1}^n x_iy_i -   \frac{(\sum_{i=1}^nx_i\sum_{i=1}^{n}y_i)}{n}$$
--
$$\hat{\beta}_1 = \frac{\sum_{i=1}^n x_iy_i -   \frac{(\sum_{i=1}^nx_i\sum_{i=1}^{n}y_i)}{n}}{(\sum_{i=1}^n x_i^2 -   \frac{(\sum_{i=1}^nx_i)^2}{n})}.$$


---
class: middle

##Properties of Estimators

- Least Square Approach provides unbiased estimators.

- Gauss Markov Theorem states that Ordinary Least Square Estimator have a lowest variance among all linear unbiased estimators.


---
class: middle


##Hypothesis Testing 

 - A statistical hypothesis is a statement about the **population parameter**.
 - Based on a sample from the population, Hypothesis Testing is a procedure used to reject or not a statistical hypothesis called the null hypothesis.


##Regression Slope Test

 - Assuming that random error follows normal distribution, we can demonstrate that
 
 $$y_i \sim N(\beta_0 + \beta_1 x_i, \sigma^2).$$
- Under $H_0$, 
 
 $$\frac{\hat{\beta}_1 - \beta_1}{\hat{\sigma^2}}\sqrt{\sum_{i=1}^n(x_i-\bar{x})^2} \sim t_{(n-2)}$$
where $n-2$ is the degree of the freedom.

 - From these results, we can test $H_0: \beta_1 = 0.$


---
class: middle

##Type of the Errors

When we performs a hypothesis test we can comete two type of errors.

- Type I error is the rejection of a true null hypothesis.
- Type II error is the non-rejection of a false null hypothesis.

Unfortunately, it is not possible to minimize both errors simultaneously, hence, we fix
a type I error and minimize Type II.

**Significance level**: the probability of rejecting the null hypothesis when it is true.

---
class: middle

##P-value

**Definition**: probability of observing the given value of the test statistic, or greater, under the null hypothesis [(FERREIRA \& PATINO)](https://www.scielo.br/scielo.php?script=sci_arttext&pid=S1806-37132015000500485).

Let's watch the videos [1](https://www.youtube.com/watch?v=HTZ8YKgD0MI) and [2](https://www.youtube.com/watch?v=RVxHlsIw_Do)  
about p-values.

I prefer the second one.

I highly recommend this [article](https://link.springer.com/content/pdf/10.1007/s10654-016-0149-3.pdf).


---
class: middle

##Example in R


```{r, warning=FALSE, message=FALSE}

mod = lm(lpsa ~lcavol, data= Prostate)
tidy(mod)

```