---
title: "Lecture 5 - Data Visualization (part II)"
subtitle: "Presentation based on R for Data Science: Written by Garrett Grolemund and Hadley Wickham"
author: " Presented by Renato Rodrigues Silva"
institute: "Federal University of Goias."
date: "(updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    


---
class: middle
##Introduction

- We introduced a ggplot2 package in the last class and have drawn a quite simple scatterplot.

- We define the `ggplot()` as function which creates a coordinate system that you can add layers to.

- Now it's time to go deeper and understand some concepts.

- What are the layers? What are they composed of?

- What are the aesthetics mapping? What are geom objects ?
What are facets? How to do that? 

- How to define a scale on the graphic?



---
class: inverse, center, middle
#Layers

---
## Layers

- Layers are all objects that we can realize on the plot. 

##Layers Components

A layer is composed of

- data and aesthetic mapping,

- a statistical transformation (stat),

- a geometric object (geom), and

- a position adjustment

---
class: inverse, center, middle
#Aesthetic mapping

---
##Aesthetic 

- Aesthetics is a visual property of the objects in the graphic.

- Aesthetics include things like the size, the shape, the x-axis, and the y-axis, or the color of the points.

- We can display a point in different ways by changing the values of its aesthetic properties. 

##Aesthetic mapping

- Aesthetic mapping is to associate an aesthetic property to the name of the variable of the data set.

-  Aesthetic mappings can be set in `ggplot()` and in individual layers.

- We will discuss layers later on.

---
#Simple Scatterplot  - mpg data 

- The aesthetics mapping was set in the layer.

```{r, eval = FALSE}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy))
```

- The aesthetics mapping was set in ggplot.  if there were more than one layer,  they would have the same aesthetics mapping.

```{r, eval = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_point()
```

In this case, both scripts result in the same plot. It is not mandatory to write the name of arguments such as `data` and `mapping`. I did because I wanted to keep things more organized.

However, when we have multiple layers might be more advantageous to do an aesthetics mapping in each layer. 



---
#Example of mpg data revisited

##Question 
Do cars with big engines use more fuel than cars with small engines? What about the "type of car"? Is the type of car influence fuel consumption?

```{r, echo=FALSE,message=FALSE,warning=FALSE}
library(tidyverse)
data(mpg)
unique(mpg$class)
```



---
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color=class)) + geom_point()
```


---
class: middle
#Example of mpg data revisited


- The colors reveal that many of the unusual points are two-seater cars. 

- In spite of they have large engines, they are more fuel-efficient than SUV.

- Sports cars have large engines like SUVs and pickup trucks, but small bodies like midsize and compact cars, which improves their gas mileage.

---
##Other options

```{r, warning=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, size=class)) + geom_point()
```

---
##Other options

```{r,warning=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, shape=class)) + geom_point()
```
---
classe: middle
#Some details

- Let's suppose for some reason that we want to draw the points with blue color.

- How Can I do that?

---
##Wrong code

```{r, eval = FALSE}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color="blue"))
```

Please, notice in that case we are not doing the aesthetics mapping, then we have to use color outside of the `aes()` function. Moreover, we have to do the the aesthetics mapping in the layer. Because the color is a property of the `geom_function()`.

##Correct code
```{r, eval = FALSE}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy),color="blue")
```



---

```{r, eval = TRUE}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color="blue"))
```

---
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy),color="blue")
```

---
class: inverse, center, middle
#Facet


---
#Facet

- We can add information about extra variables in the scatterplot using aesthetics. 

- But there is a second option (for categorical variables) that would be the use of facets. 

- Facets mean to do subplots with a subset of the data.

- The function `facet_wrap()` facet the plot by a single variable.

---

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_wrap(~ class, nrow = 2)
```

---
class; middle
###Facet 

- To facet your plot on the combination of two variables, use `facet_grid()`
- This time the formula should contain two variable names separated by a `~`.

---
```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + facet_grid(drv ~ cyl)
```

---
class: inverse, center, middle
#Geometric Objects

---
#Geometric Objects


- A **geom** is the geometrical object that a plot uses to represent data.

- ggplot2 provides over 40 geoms, and extension packages provide even more (see https://www.ggplot2-exts.org for a sampling).

- The name of the plots is associated with the type of geometric objects used. 
For example, bar charts use `geom_bar`, line charts use `geom_line`, boxplots use `geom_boxplot`; and so on. 

---

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + geom_smooth(mapping = aes(x = displ, y = hwy), se = FALSE)
```

---

```{r}
ggplot(data = mpg) + geom_histogram(mapping = aes(x = hwy)) 
```

---
```{r}
ggplot(data = mpg) + geom_bar(mapping = aes(x = class)) 
```

---
```{r}
ggplot(data = mpg) + geom_boxplot(mapping = aes(x = "", y = hwy)) 

```


---
class: inverse, center, middle
##Statistical transformations

---
##Statistical transformations

- In some situations, the raw data is enough to make a plot. For example a scatterplot in the previous slides.

- However, for some plots we need to make some statistical transformations before to make a aplot.

- For example, bar charts, histograms, and frequency polygons bin the data and then plot bin counts, the number of points that fall in each bin.

- Smoothers fit a model to the data and then plot predictions from the model.

- boxplots compute a robust summary of the distribution and then display a specially formatted box.


---
####Bar chart - Count

```{r}
ggplot(data = mpg) + geom_bar(mapping = aes(x = class)) 
```


---
####Bar chart - Proportion

```{r}
ggplot(data = mpg) + geom_bar(mapping = aes(x = class, 
y = (..count..)/sum(..count..)))
```

---
####Histogram - Density

```{r}
ggplot(data = mpg) + geom_histogram(mapping = aes(x = hwy, y = ..density..)) 
```


---
####Histogram - Density and Sturges's rule

```{r}
k = nclass.Sturges(mpg$hwy)
ggplot(data = mpg) + geom_histogram(mapping = aes(x = hwy, y = ..density..), bins = k) 
```


---
class: inverse, center, middle
##Position adjustments


---
##Position adjustments

- Position adjustments determine how to arrange
geoms that would otherwise occupy the same space

- position = "dodge".  Arrange elements side by side 

- position = "fill". Stack elements on top of one another,
normalize height 

- position = "stack". Stack elements on top of one another 

- position = "jitter". Add random noise to X and Y position
of each element to avoid overplotting

---
```{r}
ggplot(data = mpg) + geom_bar(mapping = aes(x = class, fill=drv), position="dodge") 
```


---
```{r}
ggplot(data = mpg) + geom_bar(mapping = aes(x = class, fill=drv), position="fill") 
```


---
```{r}
ggplot(data = mpg) + geom_bar(mapping = aes(x = class, fill=drv), position="stack") 
```


---

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy), position="jitter") + geom_smooth(mapping = aes(x = displ, y = hwy), se = FALSE)
```

---
class: inverse, center, middle
##Labels

---
##Labels

-  We can use the function `ggtitle()` to add a main title above the plot.

- We can use `xlab()` and `ylab()` to change the label on the x-axis and y-axis, respectively.

---

```{r}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy), position="jitter") + geom_smooth(mapping = aes(x = displ, y = hwy), se = FALSE) + ggtitle("scatterplot") +xlab("fuel-efficiency") + ylab("engines size")
```


---
class: inverse, center, middle
##Scales

---
##Scales

- Scales control the mapping from data to aesthetics. They take your data and turn it into something that you can see, like size, colour, position or shape. 

- Scales also provide the tools that let you read the plot: the axes and legends.

---
```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point(aes(colour = class)) +
scale_x_continuous(limits=c(0,10)) + scale_y_continuous(limits=c(0,50)) + scale_colour_brewer()
```

---
```{r}
ggplot(mpg ) + geom_point(aes(displ, hwy)) +
scale_x_continuous(limits=c(0,10)) + scale_y_continuous(limits=c(0,50), breaks=seq(0,44,length=5)) 
```

---
```{r}
ggplot(mpg ) + geom_point(aes(class, hwy)) +
scale_x_discrete("Cut", labels = c("2seater" = "2","compact" =  "c","midsize" = "m","minivan" = "mi", "pickup" = "pi", "subcompact" ="sb", "SUV" = "suv")) + scale_y_continuous(limits=c(0,50), breaks=seq(0,44,length=5)) 
```

---
class: inverse, center, middle
##A Complex example

---
```{r, eval = FALSE}
ggplot(mpg, aes(x = class, y = hwy, fill = drv)) +
geom_boxplot() + scale_y_continuous(name = "Fuel",breaks = seq(0,40, by=10),limits=c(0, 50)) + scale_x_discrete(name = "type of car") + ggtitle("Distribution of the hwy") + theme_bw() +
        theme(plot.title = element_text(size = 14, family = "Tahoma", face = "bold",hjust = 0.5),
              legend.text = element_text(face="bold", size=16),
              axis.title = element_text(face="bold"),
              axis.text.y=element_text(size = 11),
              axis.title.y=element_text(size = 14),
              axis.title.x =element_text(size = 14),
              axis.text.x=element_text(size = 11)) +
              scale_fill_brewer()
              
```

---
```{r, echo = FALSE}
ggplot(mpg, aes(x = class, y = hwy, fill = drv)) +
geom_boxplot() + scale_y_continuous(name = "Fuel",breaks = seq(0,40, by=10),limits=c(0, 50)) + scale_x_discrete(name = "type of car") + ggtitle("Distribution of the hwy") + theme_bw() +
        theme(plot.title = element_text(size = 14, family = "Tahoma", face = "bold",hjust = 0.5),
              legend.text = element_text(face="bold", size=16),
              axis.title = element_text(face="bold"),
              axis.text.y=element_text(size = 11),
              axis.title.y=element_text(size = 14),
              axis.title.x =element_text(size = 14),
              axis.text.x=element_text(size = 11)) +
              scale_fill_brewer()
              
```

---
class: middle

Further details can be found at

-[rstudio](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
-[book](https://ggplot2-book.org/scales.html#legends)

-[help](https://ggplot2.tidyverse.org/reference/theme.html)